High-level cognition is supported by information-rich but compressible brain activity patterns

### 2023-12-22 paper/main.tex
starting to work on network shuffling analysis requested by R2 [WIP]

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns and networks~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10,
SizeEtal18, SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11,
AdacEtal12, BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04,
DhamEtal08, KorzEtal08, BrovEtal04, LynnBass21}. Correlations or causal
associations between different brain regions simultaneously imply that
full-brain activity patterns will be compressible and also that those activity
patterns will contain redundancies. For example, the extent to which activity
patterns at one brain area can be inferred or predicted from activity patterns
at other areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region.

@@

Different brain networks also varied in how informative and compressible their
activity patterns were across experimental conditions (e.g.,
Fig.~\ref{fig:networks}). This might follow from evolutionary optimizations
that reflect the relevant constraints or demands placed on those networks. One
possibility is that cortex is organized in a hierarchy of networks ``concerned
with'' or selective to different levels of processing or function. To the
extent that different levels of processing (e.g., low-level sensory processing
versus ``deeper'' higher-level processing) reflect different stimulus
timescales~\citep[e.g.,][]{Mann20}, the network differences we observed might
also relate to the timescales at which each network is maximally
sensitive~\citep{RegeEtal18, BaldEtal17,LernEtal11, HassEtal08}.

@@

Cognitive neuroscientists are still grappling with basic questions about the
fundamental ``rules'' describing how our brains respond, and about how brain
activity patterns and the associated underlying cognitive representations and
computations are linked. We identified two aspects of brain activity patterns,
informativeness and compressibility, that appear to change systematically with
task demands and across brain networks. Our work helps to clarify how the
``neural code'' might be structured, and how the code might vary across tasks
and brain areas.

@@

>>>>

Our explorations of informativeness and compressibility are related to a much
broader literature on the correlational and causal structure of brain activity
patterns and networks~\citep{PretEtal17, OwenEtal21, RogeEtal07, RubiSpor10,
SizeEtal18, SmitEtal13b, SmitEtal13c, SrinEtal07, TomaVolk11, YeoEtal11,
AdacEtal12, BassSpor17, BullSpor09, SporHone06, SporBetz16, SporZwi04,
DhamEtal08, KorzEtal08, BrovEtal04, LynnBass21}. Correlations or causal
associations between different brain regions simultaneously imply that
full-brain activity patterns will be compressible and also that those activity
patterns will contain redundancies. For example, the extent to which activity
patterns at one brain area can be inferred or predicted from activity patterns
at other areas~\citep[e.g.,][]{OwenEtal20, ScanEtal21}, reflects overlap in the
information available in or represented by those brain areas. If brain patterns
in one area are recoverable using brain patterns in another area, then a
``signal'' used to convey the activity patterns could be compressed by removing
the recoverable activity. Predictable (and therefore redundant) brain activity
patterns are also more robust to signal corruption. For example, even if the
activity patterns at one region are unreadable or unreliable at a given moment,
that unreliability could be compensated for by other regions' activity patterns
that were predictive of the unreliable region. Whereas compressible brain
patterns a robust to spatial signal corruption, high versus low informativeness
reflects a similar (though dissociable; e.g.,
Fig.~\ref{fig:information-compression}) tradeoff between expressiveness and
robustness of \textit{temporal} patterns. Highly informative brain patterns (by
our measure; i.e., patterns that yield greater temporal decoding accuracy) are
expressive about ongoing experiences or cognitive states, since each moment's
pattern is reliably distinguishable from other moments' patterns. However, when
each moment's pattern is unique, brain activity becomes less robust to temporal
signal corruption. Our finding that brain activity patterns becomes more
informative (i.e., less robust to temporal signal corruption) and compressible
(i.e., more robust to spatial signal corruption) when cognitive engagement is
higher suggests that our brain may optimize its activity patterns to prioritize
either temporal or spatial robustness, according to task demands.

@@

We found that different brain networks varied in how informative and
compressible their activity patterns were across experimental conditions (e.g.,
Fig.~\ref{fig:networks}). This might
follow from evolutionary optimizations that reflect the relevant constraints or
demands placed on those networks. One possibility is that cortex is organized
in a hierarchy of networks ``concerned with'' or selective to different levels
of processing or function. To the extent that different levels of processing
(e.g., low-level sensory processing versus ``deeper'' higher-level processing)
reflect different stimulus timescales~\citep[e.g.,][]{Mann20}, the network
differences we observed might also relate to the timescales at which each
network is maximally sensitive~\citep{RegeEtal18, BaldEtal17,LernEtal11,
HassEtal08}.

Our reverse inference analyses (Figs.~\ref{fig:neurosynth-pca},
\ref{fig:neurosynth-summary}) also provide some insights into how neural
activity patterns change with cognitive engagement or task demands. Prior work
has shown that the components and network ``parcels'' identified through
covarying activity patterns can be highly similar even across different
tasks~\citep[including ``rest,'' e.g.,][]{SmitEtal09, LairEtal11}. We
replicated this basic finding in that the first principal components from all
four experimental conditions were strikingly similar (e.g., see the leftmost
columns of Figs.~\ref{fig:neurosynth-pca}A and~\ref{fig:neurosynth-summary}C,
D). We also found some small, though statistically reliable, systematic changes
in the weights associated with different cognitive functions across conditions
(Fig.~\ref{fig:neurosynth-summary}E). This result provided an additional way of
characterizing network-level differences across conditions
(Fig.~\ref{fig:networks}E).  Taken together, these findings suggest that although
similar networks may be involved in different tasks, the ways in which those
networks are engaged may vary systematically with task demands.

@@

Cognitive neuroscientists are still grappling with basic questions about the
fundamental ``rules'' describing how our brains respond, and about how brain
activity patterns and the associated underlying cognitive representations and
computations are linked. We identified two aspects of brain activity patterns,
informativeness and compressibility, that appear to change systematically with
task demands and across brain networks. We speculate that these changes may
reflect ongoing tradeoffs between how robust to signal corruption versus how
expressive about ongoing cognitive states our brains' activity patterns are.
Our work also provides a new framework for evaluating these tradeoffs in other
datasets, or in future studies.

@@

We recognize that ChatGPT is not omniscient, nor should it be treated as an
expert cognitive neuroscientist. We therefore reviewed ChatGPT's responses
carefully by hand to verify that they seemed reasonable to us. Whereas prior
work has often constructed such rankings by hand, we see our use of ChatGPT in
this case as a small additional ``sanity check'' on our rankings that helped us
to be slightly more objective than if we had simply created the rankings
ourselves manually.
